---
title: "Who Will Be Our President?"
author: "Franchesca Fernandez, Olivia Wang"
date: "Case Study 1: February 22, 2024"
format: 
  pdf:
    keep-tex: true
    include-in-header: 
       text: |
         \usepackage{fvextra}
         \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
         \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
    geometry: 
      - left=1in
      - right=1in
      - top=1in
      - bottom=1in
---

```{r}
#| include: false

# Loading necessary packages
library(tidyverse)
library(Sleuth2)      # the package containing the data for the case study. you may need to install this package first!
library(kableExtra)   # for creating nicely formatted tables in Quarto
library(broom)
library(performance)
library(see)
library(patchwork)
library(qqplotr)

# Loading the case study data
election <- Sleuth2::ex0825

# Creating a second dataset with Palm Beach County excluded
election_wo_pb <- election |> filter(County != "Palm Beach")

```

### Introduction

The 2000 US presidential election with forerunners Al Gore and George W. Bush was an incredibly close election -- ultimately decided the president with the votes counted in Florida. After a recount, Bush had only won Florida by around 400 votes. Democratic voters in Palm Beach County blamed the structure of the confusing voting ballot as evidence for Gore's loss, claiming accidental voting for Reform Party candidate Pat Buchanan, given his unusually high number of votes in Palm Beach. Given this data, we are interested in establishing the possible existence of the relationship of votes Bush received in relation to the votes Buchanan received, which may have been votes casted for Gore. Thus, the follow questions entail: 

-   Is there a linear relationship between the amount of votes Bush received in Florida counties and the number of votes casted for Buchanan? 

-   Can we estimate the amount of votes that Buchanan should have received in Palm Beach county, and is the actual number of votes for Buchanan substantially different?

#### Data Description

To solve these questions, we have been equipped with a data frame that contains 67 observations, where every observation is a county. There is an additional data frame where Palm Beach county has been excluded, which will be the primary data set we employ for this analysis (66 observations). Each county has corresponding information for the amount of votes Bush received (Bush2000) and the amount of votes Buchanan received (Buchanan2000).This data was sourced from the 2nd edition book titled *The Statistical Sleuth: A Course in Methods of Data Analysis*. To perform EDA, we chose to model the relationship with a scatter plot. Upon looking at this scatter plot, we notice a positive relationship. As Bush receives more voters, so does Buchanan. Generally, the counties do not stray very far from the regression line, indicating a moderate "goodness of fit". This is supported by our R2 value, which is .7518. 

```{r}
#| include: false

ggplot(data = election_wo_pb, 
       aes(x = Bush2000, y = Buchanan2000)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + labs(
    title = "2000 U.S Presidential elections in Florida counties",
    x = "Votes for George W. Bush",
    y = "Votes for Pat Buchanan",
    subtitle = "The blue line denotes the best fit line for this data"
  )
```

### Statistical Methods

In order to explore the relationship between votes for Bush and Buchanan across different counties, we assumed that the votes in Palm Beach County hold a relationship the same as any other county, excluded Palm Beach from our dataset, and accordingly built a linear regression model. The explanatory variable is the number of votes for Bush, and the response variable is the number of votes for Buchanan. The unit of observation is counties. In mathematical terms, this model can be defined as:

$$BuchananVotes = \beta_0 + \beta_1\left(BushVotes\right).$$

Let $BuchananVotes$ denote the number of votes cast for Buchanan and $BushVotes$ denote the number of votes cast for Bush.Additionally,

$\beta_0$ denotes the predicted number of votes Buchanan will receive when Bush receives no votes.

$\beta_1$ denotes the predicted change in votes for Buchanan for each additional vote for Bush.

The equation for predicting Buchanan votes from Bush votes that we computed is as follows:

$BuchananVotes$ = 65.57 + 0.00348$(BushVotes)$

With this equation, we predict that in all counties, for every additional vote for Bush, the number of votes for Buchanan increases by 0.00348 ($\beta_1$) on average, and the "baseline" is at 65.57 ($\beta_0$) votes (i.e., when Bush receives no votes, Buchanan receives 65.57 votes). The standard errors of $\beta_0$ and $\beta_1$ are 17.33 and 0.0002501, respectively. The t-value and p-value for Bush2000 are respectively 13.923 and a value less than 2*10^-16. The t-value is significantly large and the p-value is extremely small. Both of them strongly indicates that we reject the null hypothesis (i.e., no relationship between votes for Buchanan and for Bush). Therefore, we have strong evidence to conclude a relationship between the numbers of votes for the two candidates.

Our computed values can be formatted as a table for easier understanding:

```{r}
#| echo: false
#| warning: false

# Fitting the regression line 
model_eq <- lm(Buchanan2000 ~ Bush2000, data = election_wo_pb)

# Representing the regression table as a dataframe (i.e., tidying the summary() output)
model_eq_table <- summary(model_eq)$coefficients

#kable table
model_eq_table |> kbl(col.names = c("Coefficients", "Standard Error", "T-value", "P-value"), align = "c", booktabs = T, linesep="", digits = c(4, 2, 2, 5)) |> kable_classic(full_width = F, latex_options = c("HOLD_position"))
```

### Results

Additionally, we computed a 95% prediction interval for Palm Beach County to explore whether it matches the true data point. The interval we obtained is (364.709, 830.8264). We are 95% confident that the true number of votes for Buchanan in Palm Beach County falls within the calculated interval.

```{r}
#| echo: false

model_eq <- lm(Buchanan2000 ~ Bush2000, data = election_wo_pb)

prediction <- predict(model_eq, newdata = data.frame(Bush2000 = 152846), interval = "prediction", level = 0.95)
```

Examining the calculated prediction interval, we noticed that the true number of votes for Buchanan in Palm Beach is far above our prediction -- Buchanan received 3,407 votes in this county, while the upper bound of our prediction interval is 830.8. With our assumptions that Palm Beach should follow the same relationship as other counties and that some votes for Gore were miscounted as votes for Buchanan, we estimate it is likely that approximately 2,576 votes intended for Gore were cast for Buchanan. This conclusion, however, faces several limitations and assumption violations. In our regression model, we assumed linearity of the relationship, independence, normality, and equal variance of the residuals. Despite that other assumptions hold, we see the assumption of equal variance is mildly violated. In the residual plot, the variance is smaller at smaller fitted values and has a tendency to increase as fitted values increase. This indicates that the standard deviations of errors vary across different values, and tend to be larger when predicting large values.

```{r}
#| echo: false
#| message: false

# LINE assumptions check

linearity_check <- check_model(model_eq,
                               check = "linearity",
                               panel = FALSE
                               )
plot(linearity_check)[[1]]

```

```{r}
#| echo: false
#| message: false

# Equal Variance check
variance_check <- check_model(model_eq,
                              check = "homogeneity",
                              panel = FALSE
                              )
plot(variance_check)[[1]]
```

```{r}
#| echo: false
#| message: false

# Normality Check
normality_check <- check_model(model_eq,
                               check = "qq",
                               panel = FALSE
                               )
plot(normality_check)[[1]]
```

```{r}
#| echo: false

# Residual Plot
plot(model_eq, which = 1)
```

There are also limitations to take into account. First, we should note that the dataset only covers counties in Florida. We do not have enough information to draw the same conclusions for votes in other states, especially given that the political tendencies of different states vary greatly. For instance, while Florida is a Republican state, Massachusetts is a highly Democratic state. Their relationship between votes for Bush and Buchanan in 2000 could be completely different. Moreover, we should note that there are only 67 counties (i.e., 67 data points in our analysis) in Florida. This is a relatively small number -- Texas has 254 counties and Georgia has 159 counties. If more data points were taken into consideration in our analysis, the result may appear different. Given these reasons, we can only generalize our conclusions to Florida. Any bigger population will need more data collection and analysis. At the same time, our results are still highly significant. Florida, in this case, is the state that we are most interested in because it crucially determines who would be elected as the US president.

Lastly, it is important to note that we cannot guarantee any causality between the numbers of votes for Bush and for Buchanan. We do not have sufficient evidence to show that this is a cause-and-effect relationship. Therefore, the analysis only indicates the existence of a correlation between two candidates' votes.

# R Appendix

*Copy and paste all code that you used for your case study into one chunk at the end of your written report. Before submitting your case study, take one final look at the R Appendix and make sure that all code is clearly visible. If you see a line running off the side of the PDF, please split the code over multiple lines using a linebreak.*

```{r}
#| message: FALSE
#| warning: FALSE

# Loading necessary packages
library(tidyverse)
library(Sleuth2)      # the package containing the data for the case study. you may need to install this package first!
library(kableExtra)   # for creating nicely formatted tables in Quarto
library(broom)
library(performance)
library(see)
library(patchwork)
library(qqplotr)

# Loading the case study data
election <- Sleuth2::ex0825

# Creating a second dataset with Palm Beach County excluded
election_wo_pb <- election |> filter(County != "Palm Beach")

# Fitting the regression line
model_eq <- lm(Buchanan2000 ~ Bush2000, data = election_wo_pb)
# summary(model_eq)

# Representing the regression table as a dataframe (i.e., tidying the summary() output)
model_eq_table <- summary(model_eq)$coefficients

#kable table
model_eq_table |> kbl(col.names = c("Coefficients", "Standard Error", "T-value", "P-value"), align = "c", booktabs = T, linesep="", digits = c(4, 2, 2, 5)) |> kable_classic(full_width = F, latex_options = c("HOLD_position"))

# EDA

ggplot(data = election_wo_pb, 
       aes(x = Bush2000, y = Buchanan2000)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + labs(
    title = "2000 U.S Presidential elections in Florida counties",
    x = "Votes for George W. Bush",
    y = "Votes for Pat Buchanan",
    subtitle = "The blue line denotes the best fit line for this data"
  )

model_eq <- lm(Buchanan2000 ~ Bush2000, data = election_wo_pb)
summary(model_eq)

prediction <- predict(model_eq, newdata = data.frame(Bush2000 = 152846), interval = "prediction", level = 0.95)

# LINE assumptions check

linearity_check <- check_model(model_eq,
                               check = "linearity",
                               panel = FALSE
                               )
plot(linearity_check)

# Equal Variance check

variance_check <- check_model(model_eq,
                              check = "homogeneity",
                              panel = FALSE
                              )
plot(variance_check)

#Normality Check

normality_check <- check_model(model_eq,
                               check = "qq",
                               panel = FALSE
                               )
plot(normality_check, data = model_eq)

# Resid Plot
plot(model_eq, which = 1)
```
